Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-17 latex/main.tex
Updates from Overleaf

%% JK: This is copy of old Methods section "overview of sc2ts" which is basically
%% the same as this section. TODO Merge the useful text in here.

% Sc2ts (pronounced ``scoots'', optionally)
% is a method for inferring Ancestral Recombination Graphs (ARGs)
% from densely sampled genome sequence data in real time,
% in which recombination occurs at a low but significant rate.
% The basic idea is to incrementally update an ARG regularly
% with the sequences collected since the previous batch (Figure
% \ref{fig:methodology}).
% We used daily batches for SARS-CoV-2,
% and use days as the time interval for clarity in what follows.
% The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
% for each sequence in the daily batch to the current ARG.
% These copying paths will mostly consist of a new sample sequence copying
% from a node in the ARG with a small number of mutations,
% and often many samples from a daily batch will copy from the same ARG node.
% The second step is then to ``resolve'' these implied polytomies
% by using a standard neighbor-joining algorithm \cite{Saitou1987}.
% This greedy update strategy inevitably leads to unparsimonious topologies,
% and the third step is to then increase the overall parsimony of the inferred ARG
% by making some simple topological updates.
% Recombination is inferred as an integral and ongoing part of this process,
% requiring only a few additional steps to facilitate later analysis.
% The result of inference for a given day is then a genealogy recording genetic inheritance
% as well as mutation and recombination events
% for all of the samples added to the ARG up to that day.

% Sc2ts is implemented in Python building on core PyData infrastructure
% such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
% and Pandas~\cite{Mckinney2010data} ,
% using pyfaidx\cite{Shirley2015efficient} to access alignment data
% and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
% tasks.
% [TODO JK to fill this out]

@@

@@

The LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
that captures many of the key features of the joint processes of mutation and recombination.
It is widely used in a range of applications in population genetics, including ARG 
% Does Threads use LS?
inference \cite{Kelleher2019,Gunnarsson2024}.
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
The process of switching between the genomes in the reference panel is governed by a transition matrix,
and mismatches between a focal genome and a genome of the reference panel (from which it is ``copied'') are permitted as emissions.

@@

@@

\subsubsection*{Post hoc parsimony mapping}
After the main inference step we mapped the variation data for 163 sites 
in the alignments to the ARG using the Fitch-Hartigan 
algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
implemented in tskit. In contrast to the main inference process where 
deletion characters are masked as missing data, we allow for 
5 different states in these parsimony calculations: A,C,G,T and -.
By doing this, we can approximate the mutation events causing
deletions, but with the clear caveat that each site is treated 
independently. 
Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
we chose 75 sites to remap

performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.

Based on the data of Table S4 of Li2023indels

>>>>

@@

\subsubsection*{Description of sc2ts}

Sc2ts (pronounced ``scoots'', optionally)
is a method for inferring Ancestral Recombination Graphs (ARGs)
from densely sampled genome sequence data in real time,
in which recombination occurs at a low but significant rate.
The basic idea is to incrementally update an ARG regularly
with the sequences collected since the previous batch (Figure
\ref{fig:methodology}).
We used daily batches for SARS-CoV-2,
and use days as the time interval for clarity in what follows.
The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
for each sequence in the daily batch to the current ARG.
These copying paths will mostly consist of a new sample sequence copying
from a node in the ARG with a small number of mutations,
and often many samples from a daily batch will copy from the same ARG node.
The second step is then to ``resolve'' these implied polytomies
by using a standard neighbor-joining algorithm \cite{Saitou1987}.
This greedy update strategy inevitably leads to unparsimonious topologies,
and the third step is to then increase the overall parsimony of the inferred ARG
by making some simple topological updates.
Recombination is inferred as an integral and ongoing part of this process,
requiring only a few additional steps to facilitate later analysis.
The result of inference for a given day is then a genealogy recording genetic inheritance
as well as mutation and recombination events
for all of the samples added to the ARG up to that day.

Sc2ts is implemented in Python building on core PyData infrastructure
such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
and Pandas~\cite{Mckinney2010data} ,
using pyfaidx\cite{Shirley2015efficient} to access alignment data
and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
tasks.
[TODO JK to fill this out]

@@

LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
that captures many of the key features of the joint processes of mutation and recombination.
It is widely used in a range of applications in population genetics, including ARG 
% Does Threads use LS?
inference \cite{Kelleher2019,Gunnarsson2024}.
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
The process of switching between the genomes in the reference panel is governed by a transition matrix,
and mismatches between a focal genome and a genome of the reference panel (from which it is ``copied'') are permitted as emissions.

@@

\subsubsection*{Accelerating sample matching}

@@
